---
title: "Elevation_vs_Deer"
format:
  html:
    embed-resources: true
editor: visual
---

## deer_ppp vs elevation

### get data

```{r}
load("data/deer_data.Rdata")
ls()
```

```{r}
library(sf)
library(spatstat.geom)
library(readr)

# win_bc_sf <- st_as_sf(valid_window)
# win_bc <- as.owin(win_bc_sf)
elevation_im <- as.im(elevation_im)
deer_clean <- deer_ppp
```

### analysis

```{r}
library(spatstat.explore)
rho_elev <- rhohat(deer_clean, elevation_im)
plot(rho_elev, main = "Intensity of Deer vs Elevation")
```

The nonparametric estimate of deer intensity as a function of elevation reveals a sharp unimodal pattern, with significantly elevated density at very low elevations (under 300 meters) and rapidly decreasing intensity as elevation increases. This supports the hypothesis that deer occurrences are highly concentrated in lowland regions across British Columbia.

### model

```{r}
library(spatstat.geom)
library(spatstat.model)
library(spatstat.explore)  

elev_logical_im <- elevation_im
elev_logical_im$v <- !is.na(elevation_im$v)
elev_window <- as.owin(elev_logical_im)

deer_trimmed <- deer_clean[elev_window]

elevation_crop <- as.im(elevation_im, W = elev_window)

dummy_raw <- gridcentres(elev_window, nx = 100, ny = 100)

dummy_ppp <- ppp(x = dummy_raw$x,
                 y = dummy_raw$y,
                 window = elev_window)

dummy_df <- as.data.frame(dummy_ppp)
elev_values <- lookup.im(elevation_crop, dummy_df$x, dummy_df$y, naok = TRUE)

dummy_filtered <- dummy_ppp[!is.na(elev_values)]


Q <- quadscheme.logi(
  data = deer_trimmed,
  dummy = dummy_filtered,
  logical = elev_logical_im
)

model_quad <- ppm(Q, ~ elev + I(elev^2), 
                  covariates = list(elev = elevation_crop),
                  method = "logi")

summary(model_quad)



```

The fitted inhomogeneous Poisson point process model, incorporating both linear and quadratic terms of elevation, revealed a statistically significant U-shaped relationship. Intensity was minimized at mid-elevations and higher at both low and high elevations. All coefficients were highly significant (p \< 0.001), supporting the inclusion of both terms. The model used logistic approximation and achieved convergence despite some missing covariate values at a small fraction of dummy points.

```{r}
coefs <- coef(model_quad)
beta0 <- coefs["(Intercept)"]
beta1 <- coefs["elev"]
beta2 <- coefs["I(elev^2)"]

elev_seq <- seq(0, 2000, length.out = 200)

intensity <- exp(beta0 + beta1 * elev_seq + beta2 * elev_seq^2)

plot(elev_seq, intensity, type = "l", lwd = 2, col = "blue",
     xlab = "Elevation (m)", ylab = "Fitted Intensity",
     main = "Fitted Intensity vs Elevation")
```

The fitted intensity curve reveals a strong negative association between elevation and deer density. Deer are most likely to occur at lower elevations, with intensity decreasing sharply as elevation increases. This pattern suggests that elevation is a key environmental factor influencing the spatial distribution of deer, likely due to habitat preference, food availability, or accessibility at lower altitudes.

```{r}
library(viridis)
intensity_im <- predict(model_quad, eps = 2000)
plot(intensity_im,
     main = "Fitted Intensity Surface over BC",
     col = viridis(128),
     ribbon = TRUE,
     gamma = 0.5,   
     box = FALSE)

```

This fitted intensity surface shows that deer density is highest in low-elevation areas along the southern coast and river valleys of British Columbia. The intensity gradually declines across the province as elevation increases, particularly in mountainous interior regions. The spatial pattern supports the model’s inference that deer occurrences are concentrated in accessible, lowland environments.

```{r}
library(ggplot2)

coefs <- coef(model_quad)
vcov_mat <- vcov(model_quad)

elev_seq <- seq(0, 2000, length.out = 300)
X <- cbind(1, elev_seq, elev_seq^2)

log_lambda_hat <- as.vector(X %*% coefs)
se_log_lambda <- sqrt(rowSums((X %*% vcov_mat) * X))

fitted <- exp(log_lambda_hat)
upper <- exp(log_lambda_hat + 2 * se_log_lambda)
lower <- exp(log_lambda_hat - 2 * se_log_lambda)

df <- data.frame(
  Elevation = elev_seq,
  Intensity = fitted,
  Lower = lower,
  Upper = upper
)

ggplot(df, aes(x = Elevation, y = Intensity)) +
  geom_ribbon(aes(ymin = Lower, ymax = Upper), fill = "skyblue", alpha = 0.3) +
  geom_line(color = "darkblue", size = 1.2) +
  labs(
    title = "Predicted Intensity vs Elevation (with 95% CI)",
    x = "Elevation (m)",
    y = "Predicted Intensity"
  ) +
  coord_cartesian(ylim = c(0, max(upper))) +
  theme_minimal(base_size = 14)

```

The fitted intensity curve shows a steep decline with increasing elevation. Deer occurrences are most likely at lower elevations, with intensity dropping rapidly as elevation rises. Above approximately 1000 meters, the predicted intensity becomes very low and stable, suggesting that deer are largely concentrated in lower-altitude regions across British Columbia. The narrow confidence band at low elevations also indicates high certainty in this pattern.
