---
title: "Forest_vs_Deer"
author: "Shiqi Zhang"
date: "2025-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Forest Cover vs Deer in BC
### Data Preparation and Visualization
```{r}
# Load packages
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
```

```{r}
load("data/deer_data.Rdata")

# Create mask for valid overlap
valid_mask <- !is.na(forest_im)
deer_valid <- deer_ppp[valid_mask]
```
```{r}
# Plot the deer distribution and the forest cover in BC with valid window
plot(deer_valid)
plot(forest_im)
```
```{r}
# Estimate rho for forest cover
rho_forest <- rhohat(deer_valid, covariate = forest_im)

# Plot $\rho$ vs. forest cover
plot(rho_forest,
main = "Estimated rho vs. Forest Cover",
xlab = "Forest Cover (%)",
ylab = expression(hat(rho)),
xlim = c(0, 100))
```
### Fit the model
#### Baseline linear model
```{r}
# Baseline model
model_bl <- ppm(deer_valid ~ forest, covariates = list(forest = forest_im))
summary(model_bl)
```
```{r}
# Log-scaled fitted intensity surface of linear forest model
plot(model_bl,
     log = TRUE,
     se = FALSE,
     superimpose = FALSE,
     n = 300,
     main = "Log-scaled Fitted Deer Intensity – Linear Forest Cover")

# Overlay observed deer points
plot(deer_valid,
     pch = 16,
     cex = 0.4,
     cols = "black",
     add = TRUE)

```
Brighter areas indicate higher predicted intensity, mainly in regions with lower forest cover, though the pattern may oversimplify true deer distribution.
  
```{r}
# Run the quadrat test on the baseline model
quadrat_test <- quadrat.test(model_bl, nx = 4, ny = 2)
quadrat_test
```
This indicates that the linear forest cover model doesn't fully capture the complexity of the deer distribution — likely because the intensity is not linearly related to forest cover, which we already suspected based on the rhohat() curve.  
  
#### Quadratic model
```{r}
# Fit the quadratic model
model_quad <- ppm(deer_valid ~ polynom(forest, 2), covariates = list(forest = forest_im))
summary(model_quad)
```
```{r}
# Log-scaled fitted intensity surface of quadratic forest model
plot(model_quad,
     log = TRUE,
     se = FALSE,
     superimpose = FALSE,
     n = 300,
     main = "Log-scaled Fitted Deer Intensity – Quadratic Forest Cover")

# Overlay observed deer points
plot(deer_valid,
     pch = 16,
     cex = 0.4,
     cols = "black",
     add = TRUE)

```
```{r}
# Run the quadrat test on the quadratic model
quadrat_test <- quadrat.test(model_quad, nx = 4, ny = 2)
quadrat_test
```

```{r}
# Fit a higher-order polynomial
model_poly4 <- ppm(deer_ppp ~ polynom(forest, 4), covariates = list(forest = forest_im))
summary(model_poly4)
```
```{r}
# Log-scaled fitted intensity surface of quadratic forest model
plot(model_poly4,
     log = TRUE,
     se = FALSE,
     superimpose = FALSE,
     n = 300,
     main = "Log-scaled Fitted Deer Intensity – Quadratic Forest Cover")

# Overlay observed deer points
plot(deer_valid,
     pch = 16,
     cex = 0.4,
     cols = "black",
     add = TRUE)

```
```{r}
# Run the quadrat test on the quadratic model
quadrat_test <- quadrat.test(model_poly4, nx = 4, ny = 2)
quadrat_test
```

```{r}
# Extract model coefficients
coefs <- coef(model_poly4)
vcov_mat <- vcov(model_poly4)

# Create forest cover sequence
forest_seq <- seq(0, 100, length.out = 300)
X <- model.matrix(~ polynom(forest_seq, 4))

# Compute log intensity and confidence intervals
log_lambda_hat <- as.vector(X %*% coefs)
se_log_lambda <- sqrt(rowSums((X %*% vcov_mat) * X))

# Plot model prediction
plot(rho_forest, 
     main = "Model vs. rhohat: Deer Intensity vs. Forest Cover", 
     xlab = "Forest Cover (%)", ylab = expression(hat(rho)))

lines(forest_seq, exp(log_lambda_hat), col = "red", lwd = 2)
lines(forest_seq, exp(log_lambda_hat + 2 * se_log_lambda), col = "red", lty = 2)
lines(forest_seq, exp(log_lambda_hat - 2 * se_log_lambda), col = "red", lty = 2)

legend("topright", legend = c("rhohat", "Model Prediction", "±2 SE"),
       col = c("black", "red", "red"), lty = c(1,1,2), bty = "n")
```
It looks like despite the curve fitting the rho trend reasonably well, forest cover alone is not sufficient to capture where deer actually occur. The model assigns high intensity to these forests because it only "knows" forest, not geography or other limiting factors. Also, this is a non-stationary Poisson model, but it doesn’t account for spatial autocorrelation. It assumes deer only respond to forest, not elevation, population, or location-specific factors.



