---
title: "Project_Dataset"
format: 
  html:
    embed-resources: true
editor: visual
---

**Get BC Boundary**

```{r}
library(sf)
library(spatstat.geom)

load("data/BC_Parks.Rda")
win_bc_sf <- st_as_sf(DATA$Window)
win_bc <- as.owin(win_bc_sf)
plot(win_bc)
```

**Get Elevation and Forest Cover**

```{r}
elevation_im <- as.im(DATA$Elevation)
forest_im <- as.im(DATA$Forest)
plot(elevation_im)
plot(forest_im)

```

**Get population data**

```{r}
library(raster)
library(terra)

# Load population raster in WGS84
pop_raw_terra <- rast("data/can_ppp_2020_constrained.tif")  # still in WGS84

# Reproject BC to WGS84 to match population raster
bc_wgs84_sf <- st_transform(win_bc_sf, crs = crs(pop_raw_terra))
bc_wgs84_vect <- vect(bc_wgs84_sf)  # Convert sf to terra vector

# Crop using WGS84
pop_crop <- crop(pop_raw_terra, bc_wgs84_vect)  # Now they overlap

# Reproject cropped pop to BC Albers (EPSG:3005)
pop_proj <- project(pop_crop, "EPSG:3005", method = "bilinear")
```

**Read population data**

```{r}
library(terra)

bc_proj_vect <- vect(win_bc_sf)  # Back in EPSG:3005
pop_masked <- mask(pop_proj, bc_proj_vect)

# Convert forest_im to terra raster (template)
forest_r <- rast(
  nrows = nrow(forest_im$v), 
  ncols = ncol(forest_im$v),
  xmin = forest_im$xrange[1], xmax = forest_im$xrange[2],
  ymin = forest_im$yrange[1], ymax = forest_im$yrange[2],
  crs = "EPSG:3005"
)

# Resample population raster to forest resolution
pop_resampled <- resample(pop_masked, forest_r, method = "bilinear")

# Convert to matrix and then im object
vals <- terra::as.matrix(pop_resampled, wide = TRUE)
dim(vals)  # Should be [nrow, ncol]
vals_flipped <- vals[nrow(vals):1, ]

ext <- ext(pop_resampled)
xmin(ext); xmax(ext)
ymin(ext); ymax(ext)
pop_im <- im(
  vals_flipped,
  xcol = seq(xmin(ext), xmax(ext), length.out = ncol(vals_flipped)),
  yrow = seq(ymin(ext), ymax(ext), length.out = nrow(vals_flipped))
)

# check and plot
plot(pop_im, main = "Resampled Population (Aligned to Forest)")

```

```{r}
# Plot population with enhanced size and contrast
plot(pop_im,
     main = "Population Density over BC",
     ribbon = TRUE,
     col = terrain.colors(100),    # Optional: better colormap
     scale = TRUE,
     useRaster = TRUE,             # Force raster-style plotting (faster and sharp)
     box = FALSE)

# Overlay BC boundary
plot(win_bc, add = TRUE, border = "black", lwd = 1)

pop_km2 <- pop_im
pop_km2$v <- pop_km2$v * 100
plot(pop_km2,
     main = "Population Density over BC (per km²)",
     ribbon = TRUE,
     col = terrain.colors(100),
     useRaster = TRUE,
     box = FALSE)
plot(win_bc, add = TRUE, border = "black", lwd = 1)

```

Scale data point size by population

```{r}
# Convert pop_im (im) into point coordinates with values
pop_df <- as.data.frame(pop_km2)
pop_pts <- pop_df[!is.na(pop_df$value) & pop_df$value > 0, ]

# Scale dot size by population (or use fixed size)
pop_size <- sqrt(pop_pts$value) / 100  # previously was /20

# Replot
plot(win_bc, main = "Population Points over BC", border = "grey")
points(pop_pts$x, pop_pts$y,
       pch = 21,
       bg = "green",
       col = "black",
       cex = pop_size)
legend("topright",
       legend = c("1,000", "5,000", "10,000", "20,000"),
       pt.cex = sqrt(c(1000, 5000, 10000, 20000)) / 200,
       pch = 21,
       pt.bg = "green",
       col = "black",
       title = "People / km²",
       bty = "n")


```

**Get deer data**

```{r}
library(readr)
occ <- read_tsv("data/occurrence.txt", show_col_types = FALSE)
colnames(occ)
occ_bc <- occ[!is.na(occ$decimalLatitude) & !is.na(occ$decimalLongitude), ]

occ_sf <- st_as_sf(occ_bc,
                   coords = c("decimalLongitude", "decimalLatitude"),
                   crs = 4326)  # WGS84
occ_proj <- st_transform(occ_sf, crs = st_crs(win_bc_sf))

# Extract coordinates
coords <- st_coordinates(occ_proj)

library(spatstat.geom)

deer_ppp <- ppp(x = coords[, 1],
                y = coords[, 2],
                window = win_bc)

plot(deer_ppp, main = "Deer Occurrences in BC (Projected)")
```

**Output `deer_ppp`, `forest_im`, `elevation_im`, `pop_im` for further modelling**

```{r}
save(deer_ppp, forest_im, elevation_im, pop_im, file = "data/deer_data.Rdata")
```

**Reshape `elevation_im`,`forest_im`,`pop_im` to modelling with three covariates if needed.**

```{r}
# Dimensions
dim(elevation_im$v)
dim(forest_im$v)
dim(pop_im$v)

# Pixel sizes
elevation_im$xstep
forest_im$xstep
pop_im$xstep

# Bounding boxes
range(elevation_im$xcol)
range(forest_im$xcol)
range(pop_im$xcol)

```

```{r}
library(spatstat.explore)
# Resample forest and population to elevation grid
aligned <- harmonize.im(forest_im, elevation_im, pop_im)

forest_resampled    <- aligned[[1]]
elevation_resampled <- aligned[[2]]
pop_resampled <- aligned[[3]]

dim(forest_resampled$v)
dim(elevation_resampled$v)
dim(pop_resampled$v)
```

```{r}
common_mask <- !is.na(forest_resampled$v) &
               !is.na(elevation_resampled$v) &
               !is.na(pop_resampled$v)

forest_clean <- forest_resampled
forest_clean$v <- ifelse(common_mask, forest_resampled$v, NA)

elev_clean <- elevation_resampled
elev_clean$v <- ifelse(common_mask, elevation_resampled$v, NA)

pop_clean <- pop_resampled
pop_clean$v <- ifelse(common_mask, pop_resampled$v, NA)

mask_im <- im(common_mask, xcol = forest_clean$xcol, yrow = forest_clean$yrow)
valid_window <- as.owin(mask_im)

# Trim point pattern to valid area
deer_clean <- deer_ppp[valid_window]

save(deer_clean, forest_clean, elev_clean, pop_clean, valid_window,
     file = "data/deer_clean_covariates.RData")
```

**Data Visualization**

**Plot the point pattern against each covariate**

```{r}
# Standalone: Elevation only
plot(elevation_im, main = "Elevation (No Overlay)")

# Elevation + deer
plot(elevation_im, main = "Elevation")
plot(deer_ppp, add = TRUE, pch = 20, cex = 0.4, col = "green")

# Standalone: Forest only
plot(forest_im, main = "Forest Cover (No Overlay)")

# Forest + deer
plot(forest_im, main = "Forest Cover")
plot(deer_ppp, add = TRUE, pch = 20, cex = 0.4, col = "green")

# Standalone: Population only (as image)
plot(win_bc, main = "Population Points over BC", border = "grey")
points(pop_pts$x, pop_pts$y,
       pch = 21,
       bg = "blue",
       col = "blue",
       cex = pop_size)
legend("topright",
       legend = c("1,000", "5,000", "10,000", "20,000"),
       pt.cex = sqrt(c(1000, 5000, 10000, 20000)) / 200,
       pch = 21,
       pt.bg = "blue",
       col = "blue",
       title = "People / km²",
       bty = "n")

# Population + deer overlay
plot(win_bc, main = "Population Points over BC", border = "grey")
points(pop_pts$x, pop_pts$y,
       pch = 21,
       bg = "blue",
       col = "blue",
       cex = pop_size)
legend("topright",
       legend = c("1,000", "5,000", "10,000", "20,000"),
       pt.cex = sqrt(c(1000, 5000, 10000, 20000)) / 200,
       pch = 21,
       pt.bg = "blue",
       col = "blue",
       title = "People / km²",
       bty = "n")
plot(deer_ppp, add = TRUE, pch = 20, cex = 0.4, col = "green")

```

Most deer occurrences (green dots) cluster in low-elevation regions, particularly in the southern and coastal parts of BC. There are very few or no points in high-elevation areas (shown in blue-purple).

Many deer points fall into medium forest coverage areas (orange to yellow). However, not all dense forest areas are heavily occupied — there's variation.

Most deer points avoid densely populated areas (large blue circles). However, there is some overlap in sparsely populated areas, especially in the south and inland.

**Quadrat Counting (inhomogeneity)**

```{r}
# Quadrat counting: divide the window into tiles
Q <- quadratcount(deer_ppp, nx = 3, ny = 3)

# Plot quadrat counts
plot(deer_ppp, main = "Deer Occurrences with Quadrat Grid")
plot(Q, add = TRUE, col = "red")

# Chi-squared test of spatial homogeneity
quadrat.test(Q)


```

The p-value \< 2.2e-16 strongly rejects the null hypothesis of Complete Spatial Randomness (CSR). So, deer are not evenly distributed — the intensity ($\lambda$) varies across BC. This confirms that modeling intensity as a function of covariates (like elevation, forest, population) is justified.

**Nonparametric Relationship with Covariates**

```{r}
# Relationship with elevation
rho_elev <- rhohat(deer_ppp, elevation_im)
plot(rho_elev, main = "Deer Intensity vs. Elevation")

# Relationship with forest cover
rho_forest <- rhohat(deer_ppp, forest_im)
plot(rho_forest, main = "Deer Intensity vs. Forest Cover")

# Relationship with population density
rho_pop <- rhohat(deer_ppp, pop_im)
plot(rho_pop, main = "Deer Intensity vs. Population Density")

```

Deer intensity drops steeply as elevation increases. Most deer are found at low elevations, peaking around 0–200 meters. Deer prefer lowland areas, possibly due to milder climate, better forage, or habitat accessibility.

Intensity is highest at low forest cover (0–20%), with smaller secondary peaks, then declines with higher forest density. Deer might prefer edge habitats or open-forest mosaics, rather than dense forest interiors.

Very noisy plot. Some extreme values and uncertainty (wide CI bands), especially \>150 people/km². Suggests little to no clear relationship, but there's likely avoidance of highly populated areas. Most deer occur in low or sparsely populated regions.

**Kernel Smoothing for Intensity Estimation**

```{r}
# Estimate intensity using default bandwidth
lambda_hat <- density(deer_ppp)

# Plot the smoothed intensity map
plot(lambda_hat, main = "Kernel-Smoothed Deer Intensity")
plot(deer_ppp, add = TRUE, pch = 20, cex = 0.2, col = adjustcolor("green", alpha.f = 0.3))

```

High Deer Intensity in Southern BC:

The most prominent areas of high deer occurrence intensity are located in the southern part of British Columbia, particularly in the southwestern and southeastern corners. This aligns with regions where the terrain is lower in elevation, and forest cover is more extensive.

Proximity to Urban and Populated Areas:

Interestingly, some of the highest intensity zones also overlap with or are adjacent to urban centers. This suggests that deer occurrences are not entirely absent in or near developed areas — potentially due to edge habitats or the adaptability of deer to semi-urban environments.

Sparse Intensity in the North and High Elevation Zones:

The northern and interior mountainous regions show very low or zero intensity. This pattern likely reflects a combination of: Higher elevation (unsuitable habitat), Lower forest productivity, and Less human observation effort in remote regions.

Clustering Indicates Inhomogeneity:

The smoothed intensity clearly demonstrates spatial inhomogeneity in deer distribution — it is not uniform across the province. This justifies the need for modeling with spatial covariates like elevation, forest cover, and population density.
